### 2022-02-26 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_networkary_classifiers@bd34c7c24c585e30759086973bd04b941ba011bd 🚀

The accuracy paradox is the basis of a number of problems in statistical
education, and lies in the fact that, when the desired class is rare, a
model that gets less and less performant will become more and more
accurate and useful, simply by (i) underpredicting true positive cases
and (ii) over-predicting false negatives. In other words, accuracy,
defined as the proportion of predictions that are correct, is often
useless as a measure of how predictive a model is. This is particularly
true in ecological networks; the desired class (presence of an
interaction between two species) is the one we care most about, and by
far the least commmon. Herein lies the core challenge of predicting
species interactions: the extreme imbalance between classes makes the
training of predictive models difficult, and their validation even more
so as we do not reliably know which negatives are true. The connectance
(the proportion of realized interactions, usually the number of
interactions divided by the number of species pairs) of empirical
networks is usually well under 20\%, with larger networks having a lower
connectance (MacDonald et al., 2020). Recent contributions (Strydom et
al., 2021; \textbf{Becker2021OptPre?}) highlight that predictive models
of interactions can likely be improved by adding information (in the
form of, \emph{e.g.} traits), but that we do not have robust guidelines
as to how the predictive ability of these models should be evaluated,
nor about how the models should be trained. Here, by relying on simple
derivations and a series of simulations, we formulate a number of such
guidelines, specifically for the case of binary classifiers derived from
thresholded values.

Binary classifiers are usually assessed by measuring properties of their
confusion matrix, \emph{i.e.} the contingency table reporting true/false
positive/negative hits. A confusion matrix is laid out as

\[\begin{pmatrix}
    \text{tp} & \text{fp} \\
    \text{fn} & \text{tn}
\end{pmatrix} \,,\]

wherein \(\text{tp}\) is the number of interactions predicted as
positive, \(\text{tn}\) is the number of non-interactions predicted as
negative, \(\text{fp}\) is the number of non-interactions predicted as
positive, and \(\text{fn}\) is the number of interactions predicted as
negative. Almost all measures based on the confusion matrix express
rates of error or success as proportions, and therefore the values of
these components matter in a \emph{relative} way. At a coarse scale, a
classifier is \emph{accurate} when the trace of the matrix divided by
the sum of the matrix is close to 1, with other measures focusing on
different ways in which the classifer is wrong.

There is an immense diversity of measures to evaluate the performance of
classification tasks (Ferri et al., 2009). Here we will focus on five of
them with high relevance for imbalanced learning (He \& Ma, 2013); three
threshold metrics (\(\kappa\), informedness, and MMC, the Matthews
Correlation Coefficient), and two ranking metrics (the areas under the
Receiving Operator Characteristic and the Precision-Recall curves; resp.
ROC-ACU and PR-AUC). The \(\kappa\) measure of agreement (Landis \&
Koch, 1977) establishes the extent to which two observers (here the data
and the prediction) agree, and is measured as

@@

A lot of binary classifiers are built by using a regressor (whose task
is to guess the value of the interaction, and can therefore return a
value considered to be a pseudo-probability); in this case, the optimal
value below which predictions are assumed to be negative (\emph{i.e.}
the interaction does not exist) can be determined by picking a threshold
maximizing some value on the ROC curve or the PR curve. The area under
these curves (ROC-AUC and PR-AUC henceforth) give ideas on the overall
goodness of the classifier. Saito \& Rehmsmeier (2015) established that
the ROC-AUC is biased towards over-estimating performance for imbalanced
data; on the contrary, the PR-AUC is able to identify classifiers that
are less able to detect positive interactions correctly, with the
additional advantage of having a baseline value equal to prevalence.
Therefore, it is important to assess whether these two measures return
different results when applied to ecological network prediction. The ROC
curve is defined by the false positive rate on the \(x\) axis, and the
true positive rate on the \(y\) axis, and the PR curve is defined by the
true positive rate on the \(x\) axis, and the positive predictive value
on the \(y\) axis. By comparison with the previous paragraph, it is
obvious that \(F_1\) has ties to the PR curve (being close to the
expected PR-AUC), and that informedness has ties to the ROC curve
(whereby the threshold maximizing informedness is also the point of
maximal inflection on the ROC curve). One important difference between
ROC and PR is that the later does not prominently account for the size
of the true negative compartments: in short, it is more sensitive to the
correct positive predictions. In a context of strong imbalance, PR-AUC
is therefore a more stringent test of model performance.

The same approach is used to evaluate \emph{e.g.} species distribution
models (SDMs). Indeed, the training and evaluation of SDMs as binary
classifiers suffers from the same issue of low prevalence; this is not
surprising that the two fields (SDMs and network predictions) would
share methods and their attached conceptual issues, as they suffer from
data limitations, class imbalance, and the conversion of quantitative
prediction into a binary classification. In previous work, Allouche et
al. (2006) suggested that \(\kappa\) was a better test of model
performance than the True Skill Statistic (TSS; which we refer to as
Youden's informedness); these conclusions were later criticized by
Somodi et al. (2017), who emphasized that informedness' relationship to
prevalence depends on assumptions about bias in the model, and therefore
recommend the use of \(\kappa\) as a validation of classification
performance. Although this work offers recommendations about the
comparison of models, it doesn't establishes baselines or good practices
for training on imbalanced ecological data. Steen et al. (2021) show
that, when applying spatial thinning (a process that has no analogue in
networks), the best approach to train ML-based SDMs varies according to
the balancing of the dataset, and the evaluation measures used. This
suggests that there is no single ``recipe'' that is guaranteed to give
the best model. Within the context of networks, there are three specific
issues that need to be adressed. First, what values of performance
measures are we expecting for a classifier that has poor performance?
This is particularly important as it can evaluate whether low prevalence
can lull us into a false sense of predictive accuracy. Second,
independently of the question of model evaluation, is low prevalence an
issue for \emph{training}, and can we remedy it? Finally, because the
low amount of data on interaction makes a lot of imbalance correction
methods (see \emph{e.g.} Branco et al., 2015) hard to apply, which
indicators can be optimized with the least amount of positive
interaction data?

In addition to the literature on SDMs, most of the research on machine
learning application to life sciences is focused on genomics (which has
very specific challenges, see a recent discussion by Whalen et al.,
2021); this sub-field has generated largely different recommendations.
Chicco \& Jurman (2020) suggest using Matthews correlation coefficient
(MCC) over \(F_1\), as a protection against over-inflation of predicted
results; Delgado \& Tibau (2019) advocate against the use of Cohen's
\(\kappa\), again in favor of MCC, as the relative nature of \(\kappa\)
means that a worse classifier can be picked over a better one;
similarly, Boughorbel et al. (2017) recommend MCC over other measures of
performance for imbalanced data, as it has more desirable statistical
properties. More recently, Chicco et al. (2021) temper the apparent
supremacy of the MCC, by suggesting it should be replaced by Youden's
informedness (also known as \(J\), bookmaker's accuracy, and the
True-Skill Statistic) when the imbalance in the dataset may not be
representative, which is the case as species interaction networks are
often under-sampled (Jordano, 2016b, 2016a), when classifiers need to be
compared across different datasets (for example when predicting a system
in space, where undersampling varies locally; McLeod et al., 2021), and
when comparing the results to a no-skill (baseline) classifier is
important. As these conditions are likely to be met with network data,
there is a need to evaluate which measures of classification accuracy
respond in a desirable way.

\hypertarget{baseline-values}{%
\section{Baseline values}\label{baseline-values}}

@@

These two analyses point to the following recommendations: MCC is indeed
more appropriate than \(\kappa\), as although sensitive to bias, it is
sensitive in a consistent way. Informedness is appropriate at
discriminating between different skills, but confounded by bias. As both
of these measures bring valuable information on the model behavior, we
will retain them for future analyses. \(F_1\) is increasing with bias,
and should not be prioritized to evalue the performance of the model.
The discussion of sensitivity to bias should come with a domain-specific
caveat: although it is likely that interactions documented in ecological
networks are correct, a lot of non-interactions are simply unobserved;
as predictive models are used for data-inflation (\emph{i.e.} the
prediction of new interactions), it is not necessarily a bad thing in
practice to select models that predict more interactions than the
original dataset, because the original dataset misses some interactions.
Furthermore, the weight of positive interactions could be adjusted if
some information about the extent of undersampling exists (\emph{e.g.}
Branco et al., 2015). In a recent large-scale imputation of interactions
in the mammal-virus networks, Poisot et al. (2021) for example estimated
that 93\% of interactions are yet to be documented.

@@

@@

@@

@@

@@

>>>>

Ecological networks are a backbone for key ecological and evolutionary
processes; yet enumerating all of the interactions they contain is a
daunting task, as it scales with \(S^2\), \emph{i.e.} the squared
species richness TODO. Recent contributions to the field of ecological
network prediction (Strydom et al., 2021; \textbf{Becker2021OptPre?})
highlight that although interactions can be predicted by adding
ecologically relevant information (in the form of, \emph{e.g.} traits),
we do not have robust guidelines as to how the predictive ability of
these models should be evaluated, nor about how the models should be
trained. Here, by relying on simple derivations and a series of
simulations, we formulate a number of such guidelines, specifically for
the case of binary classifiers derived from thresholded values.
Specifically, we conduct an investigation of the models in terms of
their skill (ability to make the right prediction), bias (trends towards
systematically over-predicting one class), class imbalance (the relative
number of cases representing interactions), and show how these effects
interact. We conclude on the fact that models with the best
interaction-scale predictive score do not necessarily result in the most
accurate representation of the network.

The prediction of ecological interactions shares conceptual and
methodological issues with two fields in biology: species distribution
models (SDMs), and genomics. SDMs suffers from issues affecting
interactions prediction, namely low prevalence (due to sparsity of
observations/interactions) and data aggregation (due to bias in sampling
some locations/species). In previous work, Allouche et al. (2006)
suggested that \(\kappa\) was a better test of model performance than
the True Skill Statistic (TSS; which we refer to as Youden's
informedness); these conclusions were later criticized by Somodi et al.
(2017), who emphasized that informedness' is affected both by prevalence
and bias. Although this work offers recommendations about the comparison
of models, it doesn't establishes baselines or good practices for
training on imbalanced ecological data, or ways to remedy the imbalance.
Steen et al. (2021) show that, when applying spatial thinning (a process
that has no analogue in networks), the best approach to train ML-based
SDMs varies according to the balancing of the dataset, and the
evaluation measures used. This suggests that there is no single
``recipe'' that is guaranteed to give the best model. By contrast to
networks, SDMs have the advantage of being able to both thin datasets to
remove some of the sampling bias (\emph{e.g.} Inman et al., 2021), but
also to create pseudo-absences to inflate the number of supposed
negatives in the dataset (\emph{e.g.} Iturbide et al., 2015).

An immense body of research on machine learning application to life
sciences is focused on genomics (which has very specific challenges, see
a recent discussion by Whalen et al., 2021); this sub-field has
generated recommendations that do not necessarily match the current
best-practices for SDMs, and therefore hint at the importance of
domain-specific guidelines. Chicco \& Jurman (2020) suggest using
Matthews correlation coefficient (MCC) over \(F_1\), as a protection
against over-inflation of predicted results; Delgado \& Tibau (2019)
advocate against the use of Cohen's \(\kappa\), again in favor of MCC,
as the relative nature of \(\kappa\) means that a worse classifier can
be picked over a better one; similarly, Boughorbel et al. (2017)
recommend MCC over other measures of performance for imbalanced data, as
it has more desirable statistical properties. More recently, Chicco et
al. (2021) temper the apparent supremacy of the MCC, by suggesting it
should be replaced by Youden's informedness (also known as \(J\),
bookmaker's accuracy, and the True-Skill Statistic) when the imbalance
in the dataset may not be representative of the actual imbalance.

Species interaction networks are often under-sampled (Jordano, 2016b,
2016a), and this under-sampling is structured taxonomically (Beauchesne
et al., 2016), structurally (de Aguiar et al., 2019) and spatially
(Poisot, Bergeron, et al., 2021; Wood et al., 2015). As a consequence,
networks suffer from data deficiencies both within and between datasets.
This implies that the comparison of classifiers across space, when
undersampling varies locally (see \emph{e.g.} McLeod et al., 2021) is
non-trivial. Furthermore, the baseline value of classifiers performance
measures under various conditions of skill, bias, and prevalence, has to
be identified to allow researchers to evaluate whether their interaction
prediction model is indeed learning. Taken together, these
considerations highlight three specific issues for ecological networks.
First, what values of performance measures are indicative of a
classifier with no skill? This is particularly important as it can
evaluate whether low prevalence can lull us into a false sense of
predictive accuracy. Second, independently of the question of model
evaluation, is low prevalence an issue for \emph{training} or
\emph{testing}, and can we remedy it? Finally, because the low amount of
data on interaction makes a lot of imbalance correction methods (see
\emph{e.g.} Branco et al., 2015) hard to apply, which indicators can be
optimized by sacrificing least amount of positive interaction data?

It may sound counter-intuitive to care so deeply about how good a
classifier with no-skill is, as by definition, is has no skill. The
necessity of this exercise has its roots in the paradox of accuracy:
when the desired class (``two species interact'') is rare, a model that
gets less ecologically performant by only predicting the opposite class
(``these two species do not interact'') sees its accuracy increase;
because most of the guesses have ``these two species do not interact''
as a correct answer, a model that never predicts interactions would be
right an overwhelming majority of the time; it would also be utterly
useless. Herein lies the core challenge of predicting species
interactions: the extreme imbalance between classes makes the training
of predictive models difficult, and their validation even more so as we
do not reliably know which negatives are true. The connectance (the
proportion of realized interactions, usually the number of interactions
divided by the number of species pairs) of empirical networks is usually
well under 20\%, with larger networks having a lower connectance
(MacDonald et al., 2020), and therefore being increasingly difficult to
predict.

\hypertarget{a-primer-on-binary-classifier-evaluation}{%
\section{A primer on binary classifier
evaluation}\label{a-primer-on-binary-classifier-evaluation}}

Binary classifiers, which it to say, machine learning algorithms whose
answer is a categorical value, are usually assessed by measuring
properties of their confusion matrix, \emph{i.e.} the contingency table
reporting true/false positive/negative hits. A confusion matrix is laid
out as

\[\begin{pmatrix}
    \text{tp} & \text{fp} \\
    \text{fn} & \text{tn}
\end{pmatrix} \,.\]

In this matrix, tp is the number of times the model predicts an
interaction that exists in the network (true positive), fp is the number
of times the model predicts an interaction that does not exist in the
network (false positive), fn is the number of times the model fails to
predict an interaction that actually exists in the network (false
negatives), and tn is the number of times the model correctly predicts
that an interaction does not exist (true negatives). From these values,
we can derive a number of measures of model performance (see Strydom et
al., 2021 for a review of their interpretation in the context of
networks). At a coarse scale, a classifier is \emph{accurate} when the
trace of the matrix divided by the sum of the matrix is close to 1, with
other measures informing us on how the predictions fail.

There is an immense diversity of measures to evaluate the performance of
classification tasks (Ferri et al., 2009). Here we will focus on five of
them with high relevance for imbalanced learning (He \& Ma, 2013). The
choice of metrics with relevance to class-imbalanced problems is
fundamental, because as Japkowicz (2013) unambiguously concluded,
``relatively robust procedures used for unskewed data can break down
miserably when the data is skewed.'' Following Japkowicz (2013), we
focus on two ranking metrics (the areas under the Receiving Operator
Characteristic and Precision Recall curves), and three threshold metrics
(\(\kappa\), informedness, and MCC; we will briefly discuss \(F_1\) but
show early on that it has undesirable properties).

The \(\kappa\) measure (Landis \& Koch, 1977) establishes the extent to
which two observers (the network and the prediction) agree, and is
measured as

@@

A lot of binary classifiers are built by using a regressor (whose task
is to guess the value of the interaction, and can therefore return a
value considered to be a pseudo-probability); in this case, the optimal
value below which predictions are assumed to be negative (\emph{i.e.}
the interaction does not exist) can be determined by picking a threshold
maximizing some value on the ROC or the PR curve. The area under these
curves (ROC-AUC and PR-AUC henceforth) give ideas on the overall
goodness of the classifier, and the ideal threshold is the point on
these curves that minimizes the tradeoff represented in these curves.
Saito \& Rehmsmeier (2015) established that the ROC-AUC is biased
towards over-estimating performance for imbalanced data; on the
contrary, the PR-AUC is able to identify classifiers that are less able
to detect positive interactions correctly, with the additional advantage
of having a baseline value equal to prevalence. Therefore, it is
important to assess whether these two measures return different results
when applied to ecological network prediction. The ROC curve is defined
by the false positive rate on the \(x\) axis, and the true positive rate
on the \(y\) axis, and the PR curve is defined by the true positive rate
on the \(x\) axis, and the positive predictive value on the \(y\) axis.
By comparison with the previous paragraph, it is obvious that \(F_1\)
and MCC have ties to the PR curve (being close to the expected PR-AUC),
and that informedness has ties to the ROC curve (whereby the threshold
maximizing informedness is also the point of maximal inflection on the
ROC curve). One important difference between ROC and PR is that the
later does not prominently account for the size of the true negative
compartments: in short, it is more sensitive to the correct positive
predictions. In a context of strong imbalance, PR-AUC is therefore a
more stringent test of model performance.

\hypertarget{baseline-values-for-the-threshold-metrics}{%
\section{Baseline values for the threshold
metrics}\label{baseline-values-for-the-threshold-metrics}}

@@

These two analyses point to the following recommendations: MCC is indeed
more appropriate than \(\kappa\), as although sensitive to bias, it is
sensitive in a consistent way. Informedness is appropriate at
discriminating between different skills, but confounded by bias. As both
of these measures bring valuable information on the model behavior, we
will retain them for future analyses. \(F_1\) is increasing with bias,
and should not be prioritized to evalue the performance of the model.
The discussion of sensitivity to bias should come with a domain-specific
caveat: although it is likely that interactions documented in ecological
networks are correct, a lot of non-interactions are simply unobserved;
as predictive models are used for data-inflation (\emph{i.e.} the
prediction of new interactions), it is not necessarily a bad thing in
practice to select models that predict more interactions than the
original dataset, because the original dataset misses some interactions.
Furthermore, the weight of positive interactions could be adjusted if
some information about the extent of undersampling exists (\emph{e.g.}
Branco et al., 2015). In a recent large-scale imputation of interactions
in the mammal-virus networks, Poisot, Ouellet, et al. (2021) for example
estimated that 93\% of interactions are yet to be documented.

@@

\leavevmode\hypertarget{ref-Beauchesne2016ThiOut}{}%
Beauchesne, D., Desjardins-Proulx, Archambault, P., \& Gravel, D.
(2016). Thinking Outside the Boxpredicting Biotic Interactions in
Data-poor Environments. \emph{Vie Et Milieu-Life and enVironment},
\emph{66}(3-4), 333--342.

@@

\leavevmode\hypertarget{ref-deAguiar2019RevBia}{}%
de Aguiar, M. A. M., Newman, E. A., Pires, M. M., Yeakel, J. D.,
Boettiger, C., Burkle, L. A., Gravel, D., Guimarães, P. R., O'Donnell,
J. L., Poisot, T., Fortin, M.-J., \& Hembry, D. H. (2019). Revealing
biases in the sampling of ecological interaction networks. \emph{PeerJ},
\emph{7}, e7566. \url{https://doi.org/10.7717/peerj.7566}

@@

\leavevmode\hypertarget{ref-Inman2021ComSam}{}%
Inman, R., Franklin, J., Esque, T., \& Nussear, K. (2021). Comparing
sample bias correction methods for species distribution modeling using
virtual species. \emph{Ecosphere}, \emph{12}(3), e03422.
\url{https://doi.org/10.1002/ecs2.3422}

\leavevmode\hypertarget{ref-Iturbide2015FraSpe}{}%
Iturbide, M., Bedia, J., Herrera, S., del Hierro, O., Pinto, M., \&
Gutiérrez, J. M. (2015). A framework for species distribution modelling
with improved pseudo-absence generation. \emph{Ecological Modelling},
\emph{312}, 166--174.
\url{https://doi.org/10.1016/j.ecolmodel.2015.05.018}

\leavevmode\hypertarget{ref-Japkowicz2013AssMet}{}%
Japkowicz, N. (2013). Assessment Metrics for Imbalanced Learning. In
\emph{Imbalanced Learning} (pp. 187--206). John Wiley \& Sons, Ltd.
\url{https://doi.org/10.1002/9781118646106.ch8}

@@

\leavevmode\hypertarget{ref-Poisot2021GloKno}{}%
Poisot, T., Bergeron, G., Cazelles, K., Dallas, T., Gravel, D.,
MacDonald, A., Mercier, B., Violet, C., \& Vissault, S. (2021). Global
knowledge gaps in species interaction networks data. \emph{Journal of
Biogeography}, jbi.14127. \url{https://doi.org/10.1111/jbi.14127}

@@

\leavevmode\hypertarget{ref-Wood2015EffSpa}{}%
Wood, S. A., Russell, R., Hanson, D., Williams, R. J., \& Dunne, J. A.
(2015). Effects of spatial scale of sampling on food web structure.
\emph{Ecology and Evolution}, \emph{5}(17), 3769--3782.
\url{https://doi.org/10.1002/ece3.1640}
